---
title: "Reproducible Research: Peer Assessment 1"
author: "SlowLearner"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE  
    keep_md: true
---
## Introduction
This assignment is intended to demonstrate a basic understanding of Reproducible Research principles using the tools R Markdown and knitr to generate a single document artifact (as html) that includes text, code and plots created during the course of the 'research'.

### Data

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

It includes the following variables:

* `steps`: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)
* `date`: The date on which the measurement was taken in YYYY-MM-DD format
* `interval`: Identifier for the 5-minute interval when measurement was recorded

We need the following libraries:
```{r results='hide'}
library(ggplot2)
library(dplyr, quietly=TRUE)
library(stringr)
library(kableExtra)
library(cowplot)
```
### Loading and preprocessing the data
<div style="text-align: right;">[top](#TOC)</div>

No need to download the dataset. It already exists as a zip file in the working directory.
```{r}
unzip("activity.zip")
activity <- read.table("./activity.csv", sep=",", header = TRUE)
str(activity)
summary(activity)
```

## Assignment Questions

### Question 1
<div style="text-align: right;">[top](#TOC)</div>

>*"What is the mean total number of steps taken per day?"*


Group and summarize the `activity` dataset to show daily total steps. Missing values are omitted.

```{r}
daily_steps <- summarize(group_by(na.omit(activity), date), steps = sum(steps))
```

Generate a histogram that shows the frequency (number of days) of daily steps taken. Bin width is 1000.

```{r}
hist_q1 <- ggplot(daily_steps, aes(x=steps)) +
               geom_histogram(binwidth = 1000, color = "gray", fill = "wheat") +
               labs(x = "Daily total steps",
                    y = "Number of days",
                    title = "Histogram of steps per day",
                    subtitle = "Missing values excluded")
hist_q1
```

Create a results dataset that shows the calculated mean and median values of total daily steps for the `daily_steps` table.
```{r}
results_table <- data.frame(Dataset=c("NA Omitted", ""),
                            Calculation=c("Mean", "Median"),
                            Value=c(format(mean(daily_steps$steps), nsmall = 3),
                                    format(median(daily_steps$steps), nsmall = 3))
                            )
```

The table below shows the result calculations.

```{r}
results_table %>% kbl() %>% kable_material(c("striped"))
```

---

### Question 2
<div style="text-align: right;">[top](#TOC)</div>

> *"What is the average daily activity pattern?"*


Group the `activity` dataset by interval and calculate the average steps taken for each interval.
```{r}
interval_steps_mean <- summarize(group_by(na.omit(activity), interval), steps = mean(steps))
```

Generate a line chart that shows the average number of steps taken for each interval across the entire measurement period.

```{r}
ggplot(interval_steps_mean, aes(x=interval, y=steps)) +
    geom_line(linewidth = 0.5, color = "gray", linetype=1) +
    labs(x = "Interval (hhmm)",
         y = "Average steps taken",
         title = "Average daily activity pattern")
```

Find the interval that contains the maximum average number of steps taken.
```{r results="hide"}
ms <- interval_steps_mean %>% filter(steps==max(steps))
```


The maximum steps average is **`r round(ms$steps)`** at **`r str_pad(ms$interval, width=4, side="left", pad="0")`**.

---

### Question 3
<div style="text-align: right;">[top](#TOC)</div>

>*"Imputing missing values"*

Get a count of observations in the `activity` dataset that have missing values.
Checking *all* columns for missing values (making no assumptions here).
```{r results='hide'}
na_steps <- sum(is.na(activity$steps))
na_date <- sum(is.na(activity$date))
na_interval <- sum(is.na(activity$interval))
```

There are **`r na_steps`** observations with missing values in the `steps` column.<br/>
There are **`r na_date`** observations with missing values in the `date` column.<br/>
There are **`r na_interval`** observations with missing values in the `interval` column.<br/>

**Strategy for imputing missing values**

- We are only concerned with the `steps` column.  There are no missing values in any of the other columns.
- Missing values will be replaced with the corresponding (same `interval`)
`steps` average. The required lookup dataset has already been generated in Question 2 above: `interval_steps_mean`.

1. Replace missing values
   - Make a copy of the original dataset -> `nona_activity`.
   - Loop through `nona_activity`. When an observation with missing value is found, replace the missing value with the average `steps` value for the corresponding `interval`.  
   - The dataset copy now has all missing values replaced.

```{r}
nona_activity <- activity

for (i in 1:nrow(nona_activity)){
    if (is.na(nona_activity[i,1])){
        nona_activity[i,1] <- interval_steps_mean[interval_steps_mean$interval == nona_activity[i,3],2]
        }
}
```

The new dataset `nona_activity` has **`r nrow(nona_activity)`** observations
and there are **`r sum(is.na(nona_activity$steps))`** missing values.

2. Create histogram
   - Group and summarize the `nona_activity` dataset to show daily total steps.
   - Generate a histogram that shows the frequency (number of days) of daily steps taken. Bin width is 1000.

```{r}
nona_daily_steps <- summarize(group_by(nona_activity, date), steps = sum(steps))

hist_q3 <- ggplot(nona_daily_steps, aes(x=steps)) +
               geom_histogram(binwidth = 1000, color = "gray", fill = "wheat") +
               labs(x = "Daily total steps",
                    y = "Number of days",
                    title = "Histogram of steps per day",
                    subtitle = "Missing values imputed")
hist_q3
```

Append (`rbind`)the calculated mean and median values of total daily steps for the `nona_daily_steps` table onto the `results_table` created for Question 1.

```{r}
nona_results <- c("NA Imputed", "Mean", format(mean(nona_daily_steps$steps), nsmall = 3))
results_table <- rbind(results_table, nona_results)
nona_results <- c("", "Median", format(median(nona_daily_steps$steps), nsmall = 3))
results_table <- rbind(results_table, nona_results)
```

**Findings**

The table below shows the mean and median values  for both the `daily_steps`
and `nona_daily_steps` datasets.  The mean and median values are almost the same.
The only difference is that in the imputed values dataset the mean and median are identical.

```{r}
results_table %>% kbl(digits = 3) %>% kable_material(c("striped"))
```

A more significant difference is reflected in the 2 histogram y-axes (number of days).  
The NA values are not randomly scattered throughout the original `activity' dataset.
Where they are found, they always cover all intervals of any one day of observation.

Here we list the days where no observations were made.
```{r}
ds <- summarize(group_by(activity, date), steps = sum(steps))
ds[is.na(ds$steps),]
```

8 entire days have no useful observations. So when NA values are omitted (Question 1 `daily_steps`),
there are 53 days of observations.

When NA values are imputed (Question 3 `nona_daily_steps`) we end up with the full complement of
61 days of observations.

The 2 histograms are shown together below - have a look at the y-axes
```{r}
plot_grid(hist_q1, hist_q3, ncol = 2)
```

```{r}
```


---

### Question 4
<div style="text-align: right;">[top](#TOC)</div>

>*"Are there differences in activity patterns between weekdays and weekends?"*

We are using the `nona_activity` (imputed values) dataset for this question.

1. Add a new factor variable to split into weekday or weekend observations.

```{r}
nona_activity <- mutate(nona_activity, weekday=factor(
                ifelse(weekdays(as.Date(nona_activity$date)) %in% c('Saturday','Sunday'),
                                    'Weekend',
                                    'Workday'
                       )))




```

2. Create 2 datasets
   - split by the new `weekday` factor variable.
   - group by interval across the entire measurement period and calculate the `steps` average.

```{r}
work_int_steps_mean <- summarize(group_by(nona_activity[nona_activity$weekday == 'Workday',],
                                 interval), steps = mean(steps))
wend_int_steps_mean <- summarize(group_by(nona_activity[nona_activity$weekday == 'Weekend',],
                                 interval), steps = mean(steps))
```

Generate a line chart that shows the average number of steps taken for each interval across the entire measurement period.

**Method 1:** Show workday and weekend line plots separately in a single graphic.
- Generate each plot and assign to a variable.
- Use `cowplot` function `plot_grid` to arrange the plots into a single graphic.

```{r}
work_plot <- ggplot(work_int_steps_mean, aes(x=interval, y=steps)) +
             geom_line(linewidth = 0.5, color = "steelblue", linetype=1) +
             labs(x = "Interval (hhmm)",
                  y = "Average steps taken",
                  title = "Average daily activity pattern",
                  subtitle = "Workdays (Mon-Fri)")

wend_plot <- ggplot(wend_int_steps_mean, aes(x=interval, y=steps)) +
             geom_line(linewidth = 0.5, color = "coral", linetype=1) +
             labs(x = "Interval (hhmm)",
                  y = "Average steps taken",
                  title = "Average daily activity pattern",
                  subtitle = "Weekend (Sat-Sun)")

plot_grid(work_plot, wend_plot, ncol = 1)
```

**Method 2:** show workday and weekend data as a single multiline graphic. 
```{r}
ggplot() + 
    geom_line(data=work_int_steps_mean, aes(x=interval, y=steps), color='steelblue') + 
    geom_line(data=wend_int_steps_mean, aes(x=interval, y=steps), color='coral') +
    labs(x = "Interval (hhmm)",
         y = "Average steps taken",
         title = "Average daily activity pattern",
         subtitle = "Workdays and weekend") +
    scale_color_manual(name='Weekday',
                       labels=c('Workday (Mon-Fri)', 'Weekend (Sat-Sun)'),
                       values=c('steelblue', 'coral'))
```

<p style="text-align: center;">ooo0ooo</p>
